A Reproducible Data Analysis Workflow with RMarkdown, Git, Make & Docker

### 2019-10-28 rr-flow.Rmd
revision of the manuscript

---
title             : "A Reproducible Data Analysis Workflow with RMarkdown, Git, Make, & Docker"
shorttitle        : "Reproducible Data Analysis Workflow"

@@

authornote: This paper is fully reproducible using the workflow described in this paper. All materials can be found on the accompanying GitHub repository <https://github.com/aaronpeikert/reproducible-research/>.  To provide feedback or any other type of comment or questions regarding our workflow, please add an issue to the Github repository of our paper [https://github.com/aaronpeikert/reproducible-research/issues](https://github.com/aaronpeikert/reproducible-research/issues).

@@

In this tutorial, we describe a principled workflow to ensure long-term and cross-platform reproducibility of
R-based data analyses. Reproducibility means to us that scientific results together with their computational scientific workflows are reported fully, transparently and sustainably remain available, such that the workflow can be re-run by a different person and/or later in time and the results will be identical to the ones intially reported [@herouxCompatibleReproducibilityTaxonomy2018; @claerboutElectronicDocumentsGive1992]. The need to ensure reproducibility directly follows from commonly accepted rules of good scientific practice (such as the guidelines of the German Research Foundation; @dfg2019). Ensuring reproducibility is a prerequisite for replicability and a means to increase trustworthiness of empirical results [@epskamp2019rep].

Here, we combine four software tools, whose interplay can guarantuee full and sustainable computational reproducibility of data analyses and their reporting. There are various tools to enhance reproducibility [@piccoloToolsTechniquesComputational2016], of which we believe four of them are particularly important: literate programming: @ruleTenSimpleRules2019, version control @barbaHardRoadReproducibility2016, dependency management @askrenUsingMakeReproducible2016 and containerization @clyburne-sherinComputationalReproducibilityContainers2018. Importantly, only a workflow using all four in unison allows confidence in reproducing a scientific report. Even though there are various implementations of these concepts, we believe that for the R environment, the following four implementations are best suited. For literate programming, `rmarkdown`, for version control `git`, for dependency management `make` and for container `docker`. Each of them serves a meta-scientific valuable goal (reproducibility) and increases the productivity of the researcher. Each of these software solutions is highly flexible and powerful, so to master them thoroughly often years of practice are required. Luckily the researcher can learn the minimal valuable subset needed to ensure reproducibility with little practice.

We believe that for sufficient chances of long-term reproducibility, proper documentation is crucial since it answers the question "What steps have to be taken to reproduce the results?". The most precise documentation is computer code, and the most productive computer code is code that runs automatically without human interaction. These principles are what guides the use of `rmarkdown`, `git`, `make`, `docker`.

@@

This paper is fully reproducible using the workflow described in this paper. All materials, including the manuscript source and all R codes and files used to generate this manuscript, can be found on the accompanying GitHub repository <https://github.com/aaronpeikert/reproducible-research/>.

@@

The first step towards reproducibility is to have an R-script or R-markdown as primary entry point for the analysis, that runs on a local computer without error. A process, we assume, that is familiar to the reader. A next step is to make sure that all files relevant for the analysis can be moved to another computer. To this end, it is essential that all files are within one folder (or, subfolders within it) and all paths are relative to that folder. A robust solution to the problem of making sure that file access does not break across computing platforms are [Rstudio Projects](https://r4ds.had.co.nz/workflow-projects.html) and the `here` package.

@@

This folder, where everything resides that you need for analysis (code and data), is referred to as a 'project'. Working with projects is exceptionally comfortable with RStudio, an integrated development environment (IDE) for R. It is useful to organize a data-analysis project in way that strictly segregates (raw) data and code by placing them in directories called `data` and `R` [see Section 4 @marwickPackagingDataAnalytical2018]; there are also tools that automatize the standardized creation of folder structures such as `workflowr` [@workflowr].

Sometimes external requirements make it impossible that the data can be stored and shared together with the scripts. In most of the cases we have seen, these are either space constraints or privacy considerations. In these cases, of course, unrestricted reproducibility is not guaranteed. If this way of dividing data and scripts into different databases is unavoidable, we recommend that you crosscheck using checksums before loading data. To do this, a checksum must be created and filed at the time of the original analysis, and at the time of reproduction the current checksum must be compared with the stored checksum (also called a "hash", e.g., using the functions provided in package `digest`).

@@

Typically, the translation of the human-readable summary of ones computational results, be it an internal report, a presentation, or a journal manuscript, is time-consuming and error-prone. To not only create fully reproducible results but also fully reproducible reports, one needs to resort to the Literate Programming paradigm [@knuth1984literate], in which human-readable language and computer-code are interspersed to create dynamic documents whose order follow the logic of thought rather then the order of the computer. R-Markdown is a simple markup language to create dynamic documents with embedded chunks of R code that can be exported to standard formats such as documents (docx, rtf, epub), presentations (ppt, html) or websites (html) using the knitr package [@xie2015; @xie2019]. Figure 1 illustrates RMarkdown syntax and Figure 2 the resulting output. Several packages extend the scope of knitr. Of particular note are the papaja package [@papaja2018], which additional functions to enable APA-conformable document formatting, including a journal-style final typeset format, and the stargazer package [@stargazer2018], which provides journal-ready tables and reports of statistical models.

![Source of an RMarkdown](Images/rmarkdown.png)

![Rendered result](R/rmarkdown.pdf)

@@

Reproducibility at a basic level assumes that results remain identical if neither script nor data have changed. It is often not trivial to find out whether anything has changed over time and if so to "go back in time." `git` enables you to do both. A good mental model for `git` is that it takes a sequence of snapshots of all files it is supposed to track. In the language of git, these snapshots are "commits".  A commit represents a complete copy of the state of the tracked files. Each commit has a unique identifier (a hash) and a human-readable description (commit message). Going back to one state is as easy as traversing the history of all commits; it is possible to visually track changes between different versions. Finding a specific version can be made even easier by tagging snapshots, e.g., as "submission," "preprint," or "published." This collection of snapshots is called a "repository," which ideally tracks your entire R project.

@@

So to keep track of all changes on your local computer, you only need to use `git add` & `git commit` or `git commit -a` to do both. These commands need to be executed in the `terminal`, which you can access from within Rstudio (`Shift + Alt + R`).  Rstudio also offers a graphical user interface for Git and, quite likely, this interface is sufficient for almost every interaction with Git.

@@

Now one can check everything that has happened (`git log`) and look at the exact state by supplying the id of the commit or the first few digits of its hash to `git checkout`:

@@

Git also makes it especially easy to share and collaborate on a project with other researchers. A popular service for sharing via git is [Github](https://github.com). Just sharing Git repositories with the public is always free, private repositories (only visible to persons you invite) are [free for researcher](https://help.github.com/en/articles/applying-for-an-educator-or-researcher-discount). After creating a user account, you can create a new repository where Github advises you how to upload your repository from the terminal e.g., for this repository (here with user name 'aaronpeikert' and repository name 'reproducible-research'):

```{bash, eval=FALSE}
git remote add origin git@github.com:aaronpeikert/reproducible-research.git
git push -u origin master
```

@@

```{bash, eval=FALSE}
git clone git@github.com:aaronpeikert/reproducible-research.git
```

Git and Github can do even more to help you to collaborate with your fellow researchers, such as formal means to track issues and their status (open/closed/resolved) and further means to managing and merging multiple, parallel versions of code (such as branches, pull requests, merges) but this is beyond the already broad scope of this tutorial. In particular, github's issue management can be leveraged as post-publication platform to discuss manuscripts and their results [^1].
Another benefit of using Git and Github is that experimentation is highly encouraged since you can go back to any state quickly and even when you lose access to the file on your computer, everything is backed up on the Github servers.

[^1]: To comment on our paper, please add an issue to the Github repository of our paper: <https://github.com/aaronpeikert/reproducible-research/issues>.

@@

Even when you have obtained a version of the project, and you can ensure that this version is unchanged, you may not know how to reproduce the results, because it is unclear what has to be run in which order. This job gets comfortable with `make`, since it allows you to create (computational) recipes too (re-)create files.

@@

```{bash, eval=FALSE}
analysis.pdf: data/clean.csv
  Rscript -e 'rmarkdown::render("analysis.Rmd")'

@@

The first line reads: to create `analysis.pdf` the `analysis.Rmd` needs to be rendered, which depends on `data/clean.csv`. This dependency is itself a target. To create `data/clean.csv`, `R/clean.R` and `data/iris.csv` are needed. If you type `make analysis.pdf` `make` first checks whether the dependencies do exist and if not whether or it knows how to create them. So if `data/clean.csv` does not exist, `make` creates it. The same thing happens if one of the dependencies of a target is newer then the target itself, then updates everything that directly or indirectly depends on it. So if `data\iris.csv` is newer then `data\clean.csv`, make attempts to recreate `data\clean.csv` and `analysis.pdf`. If there is a dependency which is missing, and there is no recipe to make it, `make` stops with an error. It is a convention to have a target `all` which depends on everything and make it the first target in a `Makefile`, then the command `make` without any argument automatically creates everything. The button `Build All` from within RStudio also triggers this process.

```{bash, eval=FALSE}
all: analysis.pdf
```

@@

If you have followed the above workflow, a fellow researcher is three commands away from fully reproducing your analysis:

```{bash, eval=FALSE}
git clone https://github.com/sirfisher/iris.git
cd iris
make all
```

However, this relies on the crucial assumption that your computational environments are identically or sufficiently compatible, e.g., all software dependencies are installed (R and all additional R Packages) and no updates or other changes to the computational environment break or alter your analysis. As we will see shortly, this is ensuring full computational reproducibility is necessary at a deeper level than just providing the workflow itself.

@@

`docker` is a tool that allows to precisely state what computational environment an analysis relies on and recreates this environment automatically on most operating systems (Windows, OS X & Linux). `docker` does that, by setting up a virtual computer on which it executes a series of commands (e.g., installing software). It then saves the resulting state of the virtual computer in what is called an "image." This image can be started and execute commands, e.g., running an R-script or `make`. A running instance of an image is called a container. An image can be transferred and executed on any machine that has docker installed. Unregarding the machine executing the container, from the perspective of a program running inside the container, it always looks identical. There are essential differences to a traditional virtual machine, but thinking about a docker container as a computer running inside your computer is still a plausible mental model. The most important advantage over traditional virtual machines is that container are lightweight, meaning they start fast and don't need much storage. Docker achieves that by reusing large parts of the hosts operating system (how much is varying widely between Linux, OS X & Windows).

While the R programming language is considered stable and much effort is put into backward compatibility, sometimes even basic functions like `sample()`, to randomly sample from a set, change their behavior from one version to another. To ensure reproducibility in analyses based on a computers pseudo random number generator (PRNG), it is good practice to rely on fixed PRNG seeds, which are numeric values that set the PRNG into a deterministic state, i.e., the sequence of random numbers reproduces exactly.

Consider the following R-code:

@@

The usual expectation is that this code delivers the same pseudo-random five numbers regardless of the operating system or R-Version (due to `set.seed()`). Using docker, we can start an image which contains the R-Version 3.5.0, and execute the code there.

```{r random, eval=FALSE}
```

@@

```{r random, eval=FALSE}
```

```{r, child="R/random36.md"}
```

Note that this is the intended behavior and the result of a [bugfix](https://bugs.r-project.org/bugzilla/show_bug.cgi?id=17494) in the random number generator implemented from R 3.6.0 upwards.

Such change may render analyses run on previous R versions strictly not reproducible if they contain, e.g., multiple imputations, bootstrapping, simulations studies, graphics with random jitter, Bayesian estimations using sampling algorithms (such as Markov Chain Monte Carlo), or similar techniques that involve random sampling. We would like to illustrate this with a more concrete example (the full R code to reproduce this non-reproducibility is provided in the github repository of this manuscript). Using a simulated dataset with two variables `x` and `y`, we ran a linear regression model with R's `lm()` function regressing `x` on `y`. Using the `boot` package [@boot], we bootstrapped the 95% confidence intervals around the regression estimate with 1000 bootstrap samples to evaluate whether zero was inside the confindence interval. To make the analysis reproducible, we set a random seed before we bootstrapped. We ran this once in R 3.5.0:

@@

And then, we ran the identical script with identical seed in R 3.6.1:

@@

As we see from these R outputs, the confidence intervals once include zero and once do not include zero. Please note that one could discuss deeper issues about null hypothesis significance testing here but we would simply like to stress the point that computational reproducibility in the strict sense needs to capture the full computational environment.

However only rarely does an analysis depend only on base R, often a considerable number of packages is required, and they will most likely more often introduce breaking changes with updates. The whole endeavor of reproducibility is therefore at stake, every time an update is rolled out. To ensure long-term reproducibility, `docker` replicates the original computational environment of an analysis exactly. However, we do not intend to advocate that software should not be updated; updates typically promote bugfixes and provide new functionality; but full computational reproducibility is only achieved if the originally used software version(s) are documented. Among other things, this allows to trace back recent updates as to what change in what software caused a non-replication of the original results. Quite to the contrary with solutions like docker, it gets more comfortable than ever to safely update to new versions, by just bumping the R Version number of the docker image.

This convenience is possible to the efforts of the [rocker project](https://github.com/rocker-org/rocker) [@boettigerIntroductionRockerDocker2017a] which provides docker images preconfigured with an R-installation of a fixed R-version. These rely on MRAN [@revolutionanalyticsReproducibilityUsingFixed2019], a repository for R-Packages, which is fixed to the last date the R-version of the image was the most recent. Building upon these rocker-images researchers can build there own docker-images with the required R-packages. The rocker project also provides images which do include RStudio (`rocker/rstudio`), the `tidyverse` package (`rocker/tidyverse`) and the `rmarkdown` package with Latex (`rocker/verse`). Since the describes workflow relies on `rmarkdown` we suggest using the `rocker/verse` image (which also contains `rstudio` & `tidyverse`). These images are stored on [dockerhub](https://hub.docker.com) [@dockerinc.DockerHub2019].

@@

```{bash, eval=FALSE}
FROM rocker/verse:3.6.1
RUN install2.r --error --skipinstalled\
  pacman here pander
RUN installGithub.r\
  crsh/papaja benmarwick/wordcountaddin
WORKDIR /home/rstudio
```

@@

rr-flow.pdf: rr-flow.Rmd reproducible-research.bib
	$(run) Rscript -e 'rmarkdown::render("$(current_dir)/$<")'
```

@@

This workflow directly relies on three software packages: `git`, `make` and `docker`. Of course it depends also indirectly on R, RStudio & RMarkdown, but they do not need to be installed. The installation process depends on the operating system, e.g. on OS X, make is allways availible or on Linux git & make are often allready installed. In the following section we share what have experienced as the easiest options, however in the past how docker should be installed has changed several times, so when in doubt search the internet with: `how to install docker on [your os]`.

@@

## OS X

As make already ships with OS X, you only need `git` and `docker-desktop`. We suggest using the package manager `homebrew`, which you can install from [https://docs.brew.sh/Installation](https://docs.brew.sh/Installation), to install `docker` (git will be installed during the installation of `homebrew`):

@@

We provide a reproducible analysis as working example via [github](https://github.com/aaronpeikert/workflow-showcase) as a testbed for users. The example shows a minimalistic analysis of the 'Considerations of Future Consequences (CFC) Scale'. The analysis demonstrates a succesful implementation of our workflow including downloading external data, comparing their state using a checksum, and running a confirmatory factor analysis on the first few items using lavaan [@lavaan2012]. Once all mentioned tools are installed on a computer, the following few steps are necessary to reproduce our analysis:

```{bash, eval=FALSE}
git clone https://github.com/aaronpeikert/workflow-showcase.git
cd reproducible-research
make build
make all DOCKER=TRUE
```

@@

```{bash, eval=FALSE}
git clone https://github.com/aaronpeikert/reproducible-research.git
cd reproducible-research
make build
make all DOCKER=TRUE
```

@@

Each tool reduces the number of ways things can go wrong. Dynamic reporting with `RMarkdown` guarantuees consistency between computational results and their reporting; Version control with `git` guarantuees permanence and consistency across multiple versions of data and code; Dependency management with `make` guarantuees defined entry-points and dependency resolution; containerization with `docker` guarantuees reproducibility of the full computational environment. This tightening of loose ends, we believe, does not restrict researchers but enables them to operate on a solid basis to deliver sound research.

@@

While this approach here was designed to scale well with the complexity of a computing-intense Project, we realise that this flexibility comes with a steep learning curve. A few packages are easing this burden for different parts of the workflow when this flexibility is not needed. The use of `rmarkdown` within a project, tracked with `git` can be simplified with the [workflowr-package](https://github.com/jdblischak/workflowr) [@workflowr]. The [drake-package](https://github.com/ropensci/drake) [@drake] is directly inspired by `make` but does only require knowledge in R, but can only handle dependencies within R. The [liftr-package](https://cran.r-project.org/web/packages/liftr/vignettes/liftr-intro.html) [@liftr] and the [holepunch-package](https://karthik.github.io/holepunch/) [@holepunch] automize the use of docker. The former is perfectly compatible with the described workflow and we recommend it to users uncomftable with docker. The later uses [binder](https://mybinder.org) to move the analysis to the cloud so that no local installation of docker is required. holepunch is well suited for simple analysis with low computational demands, because of memory and computing time limitations of binder. There are several alternatives to docker which fix the dependencies on R-packages. [packrat](https://rstudio.github.io/packrat/) [@packrat] is a way to fix package version via local installations of packages in the project, however it does not garuantees more then a fixed set of packages (Not R-version or system dependencies). Similar approaches are taken by [jetpack](https://github.com/ankane/jetpack) [@jetpack], [miniCRAN](https://cran.r-project.org/web/packages/miniCRAN/vignettes/miniCRAN-introduction.html) [@miniCRAN] and [checkpoint](https://cran.r-project.org/web/packages/checkpoint/vignettes/checkpoint.html) [@checkpoint]. Notable is also the package [`reprex`](https://github.com/tidyverse/reprex) [**repr**oducible **ex**ample, @reprex], but its scope is limited to small examples.

>>>>

---
title             : "A Reproducible Data Analysis Workflow with R Markdown, Git, Make, & Docker"
shorttitle        : "Reproducible Data Analysis Workflow"

@@

authornote: This paper is fully reproducible using the workflow described in this paper. All materials can be found on the accompanying GitHub repository <https://GitHub.com/aaronpeikert/reproducible-research/>.  To provide feedback or any other type of comment or questions regarding our workflow, please add an issue to the GitHub repository of our paper [https://GitHub.com/aaronpeikert/reproducible-research/issues](https://GitHub.com/aaronpeikert/reproducible-research/issues).

@@

In this tutorial, we describe a principled workflow to ensure long-term and cross-platform reproducibility of
R-based data analyses. Reproducibility means to us that statistica results together with their computational workflows are reported fully, transparently and sustainably remain available, such that the workflow can be re-run by a different person and/or later in time and the results will be identical to the ones intially reported [@herouxCompatibleReproducibilityTaxonomy2018; @claerboutElectronicDocumentsGive1992]. The need to ensure reproducibility directly follows from commonly accepted rules of good scientific practice (such as the guidelines of the German Research Foundation; @dfg2019). Ensuring reproducibility is a prerequisite for replicability and a means to increase trustworthiness of empirical results [@epskamp2019rep].

Here, we combine four software tools, whose interplay can guarantuee full and sustainable computational reproducibility of data analyses and their reporting. There are various tools to enhance reproducibility [@piccoloToolsTechniquesComputational2016], of which we believe four of them are particularly important: literate programming: [@ruleTenSimpleRules2019], version control [@barbaHardRoadReproducibility2016], dependency management @[askrenUsingMakeReproducible2016] and containerization [@clyburne-sherinComputationalReproducibilityContainers2018]. Importantly, only a workflow using all four concepts in unison guarantuee confidence in reproducing a scientific report. Even though there are various implementations of these concepts, we believe that for the R environment <CITATION NEEDED>, the following four implementations are best suited. For literate programming, R Markdown <CITATION NEEDED>, for version control Git <CITATION NEEDED>, for dependency management Make <CITATION NEEDED>, and for container Docker <CITATION NEEDED>. Each of them serves a meta-scientific valuable goal (reproducibility) and increases the productivity of the researcher. Each of these software solutions is highly flexible and powerful, so to master them fully, a considerable amount of practice is necessary. However, for our purposes, it is sufficient to master a minimal valuable subset of functions to ensure reproducibility.

To maximize chances of long-term reproducibility, proper documentation of all steps taken and their context is crucial since it answers the question "What steps have to be taken to reproduce the results?". The most precise documentation is computer code, and the most productive computer code is code that runs automatically without human interaction. This principle is what guided our choice to use of R Markdown, Git, Make, and Docker.

@@

This paper is fully reproducible using the workflow described in this paper. All materials, including the manuscript source and all R codes and files used to generate this manuscript, can be found on the accompanying GitHub repository <https://GitHub.com/aaronpeikert/reproducible-research/>.

@@

The first step towards reproducibility is to have an R script or R Markdown file as primary entry point for the analysis, which runs on a local computer without error. Next, one has to make sure that all files relevant for the analysis can be moved to another computer. To this end, it is essential that all files are within one folder (or, subfolders within it) and all paths are relative to that folder. A robust solution to the problem of making sure that file access does not break across computing platforms are [Rstudio projects](https://r4ds.had.co.nz/workflow-projects.html) and the `here` <CITATION NEEDED> package.

@@

The folder, in which all fies reside that you need for analysis (code and data), is referred to as a 'project'. Working with projects is particularly comfortable with RStudio, an integrated development environment (IDE) for R. It is useful to organize a data-analysis project in way that strictly segregates (raw) data and code by placing them in directories called `data` and `R` [see Section 4 in @marwickPackagingDataAnalytical2018]; there are also tools that automatize the standardized creation of folder structures such as `workflowr` [@workflowr].

Sometimes external requirements make it impossible that the data can be stored and shared together with the scripts. In most of the cases we have seen, these are either space constraints or privacy considerations. In these cases, of course, unrestricted reproducibility is not guaranteed. If this way of dividing data and scripts into different databases is unavoidable, we recommend to validate all data files using checksums (also called a "hash", e.g., using the functions provided in package `digest` <CITATION NEEDED>) before analyzing them. To do this, a checksum must be created and stored at the time of the original analysis. At the time of reproduction, the current checksum must be compared with the stored checksum.

@@

The translation of computational results into a human-readable summary, for example into a technical report, a presentation, or a journal manuscript, is time-consuming and error-prone. Typical errors result from copy\&paste mistakes, erroneous rounding, or missed updates of the manuscript when the computer coder has changed. To not only create fully reproducible results but also fully reproducible reports, one needs to resort to the Literate Programming paradigm [@knuth1984literate], in which human-readable language and computer-code are interspersed to create dynamic documents whose order follows the logic of thought rather then the order of the computer. R Markdown is a simple markup language to create dynamic documents with embedded chunks of R code that can be exported to standard formats such as documents (docx, rtf, epub), presentations (ppt, html) or websites (html) using the knitr package [@xie2015; @xie2019]. Figure 1 illustrates R Markdown syntax and Figure 2 the resulting output. Several packages extend the functionality of knitr. Of particular note are the `papaja` package [@papaja2018], which additional functions to enable APA-conformable document formatting, including a journal-style final typeset format, and the `stargazer` package [@stargazer2018], which provides journal-ready tables and reports of statistical models. Here is a simple example of how `papaja` can be used to create dynamic reports in APA style:

![Source of an R Markdown](Images/R Markdown.png)

![Rendered result](R/R Markdown.pdf)

@@

Reproducibility at a basic level assumes that results remain identical if neither script nor data have changed. It is often not trivial to find out whether anything has changed over time and if so to "go back in time." `git` enables you to do both. A good mental model for `git` is that it takes a sequence of snapshots of all files it is supposed to track. In the language of git, these snapshots are "commits".  A commit represents a complete copy of the state of all tracked files. Each commit has a unique identifier (a hash) and a human-readable description (commit message). Going back to one state is as easy as traversing the history of all commits; it is possible to visually track changes between different versions. Finding a specific version can be made even easier by tagging snapshots, e.g., as "submission," "preprint," or "published." This collection of snapshots is called a "repository," which ideally tracks your entire R project.

@@

To keep track of all changes on your local computer, one only needs to use `git add` & `git commit` or `git commit -a` to do both. These commands need to be executed in the `terminal`, which one can access from within Rstudio (`Shift + Alt + R`).  Rstudio also offers a graphical user interface for Git and, quite likely, this interface is sufficient for almost every interaction with Git.

@@

Now one can inspect all changes (`git log`) and examine any previous state by stating the identifier of the commit to `git checkout`:

@@

Git also makes it especially easy to share and collaborate on a project with other researchers. A popular service for sharing materials via Git is [GitHub](https://GitHub.com). Just sharing Git repositories with the public is always free, private repositories (only visible to persons you invite) are [free for researchers](https://help.GitHub.com/en/articles/applying-for-an-educator-or-researcher-discount). After creating a user account, one can create a new repository where GitHub advises you how to upload your repository from the terminal, e.g., for our repository (here with user name 'aaronpeikert' and repository name 'reproducible-research'):

```{bash, eval=FALSE}
git remote add origin git@GitHub.com:aaronpeikert/reproducible-research.git
git push -u origin master
```

@@

```{bash, eval=FALSE}
git clone git@GitHub.com:aaronpeikert/reproducible-research.git
```

Git and GitHub can do even more to help you to collaborate with your fellow researchers, such as formal means to track issues and their status (open/closed/resolved) and further means to managing and merging multiple, parallel versions of code (such as branches, pull requests, merges) but this is beyond the scope of this tutorial. In particular, GitHub's issue management can be leveraged as post-publication platform to discuss manuscripts and their results [^1].
Another benefit of using Git and GitHub is that experimentation is highly encouraged since you can go back to any state quickly and even when you lose access to the file on your computer, everything is backed up on the GitHub servers.

[^1]: To comment on our paper, please add an issue to the GitHub repository of our paper: <https://GitHub.com/aaronpeikert/reproducible-research/issues>.

@@

Even when you have obtained a given version of a project to reproduce, and you can ensure that this version is unchanged, you may not know how to reproduce the results, because it may be unclear what files have to be run in which order. This is particularly the case when complex preprocessing pipelines are part of the computation and/or external dependencies are many. Handling such dependencies is comfortable with Make because it allows you to create (computational) recipes to (re-)create files.

@@

```{bash, eval=FALSE}
analysis.pdf: data/clean.csv
  Rscript -e 'R Markdown::render("analysis.Rmd")'

@@

The first line reads: to create `analysis.pdf` the `analysis.Rmd` needs to be rendered, which depends on `data/clean.csv`. This dependency is itself a target. To create `data/clean.csv`, `R/clean.R` and `data/iris.csv` are needed. If you type `make analysis.pdf`, `make` first checks whether the dependencies do exist and, if not, creates them. Here, if `data/clean.csv` does not exist, `make` creates it. Also, if one of the dependencies of a target is newer then the target itself, then Make updates everything that directly or indirectly depends on the target. Here, if `data\iris.csv` is newer then `data\clean.csv`, Make attempts to recreate `data\clean.csv` and `analysis.pdf`. If there is a dependency which is missing, and there is no recipe to make it, `make` stops with an error. It is a convention to have a target `all` which depends on everything and make it the first target in a `Makefile`, then the command `make` without any argument automatically creates everything. The button `Build All` from within RStudio triggers this process.

@@

If you have followed our workflow as presented thus far, a fellow researcher is only three commands away from fully reproducing your analysis:

```{bash, eval=FALSE}
git clone https://GitHub.com/sirfisher/iris.git
cd iris
make all
```

However, succesful reproducibility relies on the crucial assumption that your computational environments are identically or sufficiently compatible, e.g., all software dependencies are installed (R and all additional R Packages) and no updates or other changes to the computational environment break or alter your analysis. As we will shortly see, ensuring full computational reproducibility requires one further level of documentation, that is, documentation of the computational context.

@@

Docker is a tool that allows to encapsulate, share, and re-create a computational environment on most operating systems (Windows, macOS & Linux). Docker achieves these goals by setting up a virtual computer on which it can execute well-defined series of commands (e.g., installing software). It then saves the resulting state of the virtual computer in what is called an "image." This image can be started and execute commands, e.g., running an R-script or `make`. A running instance of an image is called a container. An image can be transferred and executed on any machine that has docker installed. Irrespetictive of the machine that is executing the container, the computational environment is the same for the programs running inside the container. The most important advantage over traditional virtual machines is that containers are lightweight, that is, they start fast and don't need much storage space. Docker achieves this by reusing large parts of the hosts operating system (how much is varying widely between Linux, macOS & Windows).

With the following example, we demonstrate why it is important to store the computational environment. Generally, with containers we would like to safe-guard against changes to the computational environment with unexpected changes, e.g., changes in the functionality or default options in packages or even in the R environment itself. While the R programming language is considered stable and much effort is put into backward compatibility, sometimes even basic functions like `sample()`, to randomly sample from a set, change their behavior from one version to another. To ensure reproducibility in analyses based on a computers pseudo random number generator (PRNG), it is good practice to rely on fixed PRNG seeds, which are numeric values that set the PRNG into a deterministic state, i.e., the sequence of random numbers reproduces exactly.

Consider the following R-code to randomly draw five numbers between 1 and 10:

@@

The usual expectation is that this code delivers the same pseudo-random five numbers regardless of the operating system or R-Version (due to `set.seed()`). Using Docker, we can start an image which contains the R-Version 3.5.0, and execute the code there.

```{r random, eval=FALSE}
```
This outputs:

@@

```{r random, eval=FALSE}
```
This outputs:
```{r, child="R/random36.md"}
```

Note that this is the intended behavior and the result of a [bugfix](https://bugs.r-project.org/bugzilla/show_bug.cgi?id=17494) in the random number generator implemented from R 3.6.0 upwards. Such changes may render analyses run on previous R versions strictly not reproducible if they contain, e.g., multiple imputations, bootstrapping, simulations studies, graphics with random jitter, Bayesian estimations using sampling algorithms (such as Markov Chain Monte Carlo), or similar techniques that involve random sampling. We would like to illustrate this with a more concrete example (the full R code to reproduce this non-reproducibility is provided in the GitHub repository of this manuscript). Using a simulated dataset with two variables `x` and `y`, we ran a linear regression model with R's `lm()` function regressing `x` on `y`. Using the `boot` package [@boot], we bootstrapped the 95% confidence intervals around the regression estimate with 1000 bootstrap samples to evaluate whether zero was inside the confindence interval. To make the analysis reproducible, we set a random seed. We ran this once in R 3.5.0:

@@

Then, we ran the identical script with identical seed in R 3.6.1:

@@

As we see from these R outputs, the confidence intervals once include zero and once do not include zero. Please note that one could discuss deeper issues about null hypothesis significance testing here but we would simply like to stress the point that computational reproducibility in the strict sense requires capturing the full computational environment.

Only rarely does an analysis depend only on base R. Often, a considerable number of packages is required, and they will increase the likelihood of breaking reproducibiity with every update to any of the packages of base R. The whole endeavor of reproducibility is therefore at stake, every time an update is rolled out. To ensure long-term reproducibility, `docker` replicates the original computational environment of an analysis exactly. Note that we do not intend to advocate that software should not be updated; updates typically promote bugfixes and provide new functionality; but full computational reproducibility is only achieved if the originally used software versions are documented. Among other things, this allows to trace back recent updates as to what change in what package caused a non-reproduction of the original results. Quite to the contrary, with solutions like docker, it gets more comfortable than ever to safely update to new versions, by just bumping the R Version number of the docker image.

This convenience is possible to the efforts of the [rocker project](https://GitHub.com/rocker-org/rocker) [@boettigerIntroductionRockerDocker2017a] which provides docker images preconfigured with an R-installation of a fixed R-version. These rely on MRAN [@revolutionanalyticsReproducibilityUsingFixed2019], a repository for R-Packages, which is fixed to the last date the R-version of the image was the most recent. Building upon these rocker-images researchers can build there own docker-images with the required R-packages. The rocker project also provides images which do include RStudio (`rocker/rstudio`), the `tidyverse` package (`rocker/tidyverse`) and the `R Markdown` package with Latex (`rocker/verse`). Since the describes workflow relies on `R Markdown` we suggest using the `rocker/verse` image (which also contains `rstudio` & `tidyverse`). These images are stored on [dockerhub](https://hub.docker.com) [@dockerinc.DockerHub2019].

@@

```{bash, eval=FALSE}
FROM rocker/verse:3.6.1
RUN install2.r --error --skipinstalled\
  pacman here pander
RUN installGitHub.r\
  crsh/papaja benmarwick/wordcountaddin
WORKDIR /home/rstudio
```

@@

rr-flow.pdf: rr-flow.Rmd reproducible-research.bib
	$(run) Rscript -e 'R Markdown::render("$(current_dir)/$<")'
```

@@

This workflow directly relies on three software packages: `git`, `make` and `docker`. Of course it depends also indirectly on R, RStudio & R Markdown, but they do not need to be installed. The installation process depends on the operating system, e.g. on macOS, make is allways availible or on Linux git & make are often allready installed. In the following section we share what have experienced as the easiest options, however in the past how docker should be installed has changed several times, so when in doubt search the internet with: `how to install docker on [your os]`.

@@

## macOS

As make already ships with macOS, you only need `git` and `docker-desktop`. We suggest using the package manager `homebrew`, which you can install from [https://docs.brew.sh/Installation](https://docs.brew.sh/Installation), to install `docker` (git will be installed during the installation of `homebrew`):

@@

We provide a reproducible analysis as working example via [GitHub](https://GitHub.com/aaronpeikert/workflow-showcase) as a testbed for users. The example shows a minimalistic analysis of the 'Considerations of Future Consequences (CFC) Scale'. The analysis demonstrates a succesful implementation of our workflow including downloading external data, comparing their state using a checksum, and running a confirmatory factor analysis on the first few items using lavaan [@lavaan2012]. Once all mentioned tools are installed on a computer, the following few steps are necessary to reproduce our analysis:

```{bash, eval=FALSE}
git clone https://GitHub.com/aaronpeikert/workflow-showcase.git
cd reproducible-research
make build
make all DOCKER=TRUE
```

@@

```{bash, eval=FALSE}
git clone https://GitHub.com/aaronpeikert/reproducible-research.git
cd reproducible-research
make build
make all DOCKER=TRUE
```

@@

Each tool reduces the number of ways things can go wrong. Dynamic reporting with `R Markdown` guarantuees consistency between computational results and their reporting; Version control with `git` guarantuees permanence and consistency across multiple versions of data and code; Dependency management with `make` guarantuees defined entry-points and dependency resolution; containerization with `docker` guarantuees reproducibility of the full computational environment. This tightening of loose ends, we believe, does not restrict researchers but enables them to operate on a solid basis to deliver sound research.

@@

While this approach here was designed to scale well with the complexity of a computing-intense Project, we realise that this flexibility comes with a steep learning curve. A few packages are easing this burden for different parts of the workflow when this flexibility is not needed. The use of `R Markdown` within a project, tracked with `git` can be simplified with the [workflowr-package](https://GitHub.com/jdblischak/workflowr) [@workflowr]. The [drake-package](https://GitHub.com/ropensci/drake) [@drake] is directly inspired by `make` but does only require knowledge in R, but can only handle dependencies within R. The [liftr-package](https://cran.r-project.org/web/packages/liftr/vignettes/liftr-intro.html) [@liftr] and the [holepunch-package](https://karthik.GitHub.io/holepunch/) [@holepunch] automize the use of docker. The former is perfectly compatible with the described workflow and we recommend it to users uncomftable with docker. The later uses [binder](https://mybinder.org) to move the analysis to the cloud so that no local installation of docker is required. holepunch is well suited for simple analysis with low computational demands, because of memory and computing time limitations of binder. There are several alternatives to docker which fix the dependencies on R-packages. [packrat](https://rstudio.GitHub.io/packrat/) [@packrat] is a way to fix package version via local installations of packages in the project, however it does not garuantees more then a fixed set of packages (Not R-version or system dependencies). Similar approaches are taken by [jetpack](https://GitHub.com/ankane/jetpack) [@jetpack], [miniCRAN](https://cran.r-project.org/web/packages/miniCRAN/vignettes/miniCRAN-introduction.html) [@miniCRAN] and [checkpoint](https://cran.r-project.org/web/packages/checkpoint/vignettes/checkpoint.html) [@checkpoint]. Notable is also the package [`reprex`](https://GitHub.com/tidyverse/reprex) [**repr**oducible **ex**ample, @reprex], but its scope is limited to small examples.
